---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
    knitr:
        opts_knitr:
            verbose: TRUE
        opts_chunk:
            echo: TRUE
---


## Loading and preprocessing the data
Loading and examining the data
```{r readdata, eval = TRUE, cache = TRUE}
activity <- read.csv("activity.csv")
str(activity)
```
Changing the date column to date format in R and summing data into days
```{r formatdata, eval = TRUE, cache = TRUE}
activity$date <- as.Date(activity$date)
byday <- aggregate(steps ~ date, activity, sum)
str(byday)
```
Making a histogram of the total number of steps per day
```{r histogram, eval = TRUE, cache = TRUE}
hist(byday$steps, xlab = "Number of Steps", main = "Steps per day", col = "blue")
```

## What is mean total number of steps taken per day?
Calculating the mean by day
```{r mean, eval=TRUE}
stepmean <- round(mean(byday$steps, na.rm=TRUE))
```
Calculating the median
```{r median, eval=TRUE}
stepmedian <- median(byday$steps, na.rm=TRUE)
```
<li> the mean is ```r stepmean```
<li> the median is ```r stepmedian```

## What is the average daily activity pattern?
```{r by_interval, eval = TRUE, cache=TRUE}
byinterval <- aggregate(steps ~ interval, activity, mean)
plot(byinterval$interval, byinterval$steps, type="l", lwd=2, xlab="interval", ylab="steps", main="Average Daily Activity Pattern")
intmax <- which.max(byinterval$steps)
intmax <- byinterval[intmax,1]
```
<li> The 5-minute interval with the maximum number of steps is ```r intmax```

## Imputing missing values
Calculating total NAs
```{r imputing0, cache=TRUE, eval=TRUE}
sum(is.na(activity[,1]))
```

Starting by inputting 0 in for NA's on 2012-10-01
```{r imputing1, cache=TRUE, eval=TRUE}
NAvect1 <- which(is.na(activity[,1])==TRUE & activity[,2]=="2012-10-01")
imputed <- activity
imputed[NAvect1,1] <- 0
```

Using Impute function from the Hmisc package
```{r imputing2, cache=TRUE}
library(Hmisc)
imputed$steps <- impute(imputed$steps, fun=mean)
```

Finally, lets check to make sure we got rid of those NA's
```{r imputing3, cache=TRUE, eval=TRUE}
summary(imputed)
```

Now lets analyze and compare our imputed data
```{r histogram2, eval = TRUE}
imputedbyday <- aggregate(steps ~ date, imputed, sum)
par(mfrow = c(1,2))
hist(imputedbyday$steps, xlab = "Number of Steps", main = "Steps per day (imputed)", col = "red")
hist(byday$steps, xlab = "Number of Steps", main = "Steps per day (original)", col = "blue")
impstepmean <- round(mean(imputedbyday$steps))
impstepmedian <- median(imputedbyday$steps)
```
<li>Original mean =```r stepmean``` Imputed mean=```r impstepmean```
<li>Original median = ```r stepmedian``` Imputed median=```r impstepmedian```

## Are there differences in activity patterns between weekdays and weekends?

looping through data to create weekday/weekend factor
```{r daytypes, eval = TRUE}
imputed$weekday <- as.character(rep(NA))
imputed$date <- as.POSIXct(imputed$date)
imputed$weekday <- weekdays(imputed$date)
for (i in 1:nrow(imputed)){
    if (imputed[i,4]=="Monday" | imputed[i,4]=="Tuesday" | imputed[i,4]=="Wednesday" | imputed[i,4]=="Thursday" | imputed[i,4]=="Friday"){
    imputed[i,4] <- "weekday"
    } else {
    imputed[i,4] <- "weekend"
    }
}
imputed$weekday <- as.factor(imputed$weekday)
```
Aggregating interval means by daytype
```{r impmeans3, eval=TRUE}
imputedmeans <- aggregate(steps ~ interval + weekday, data = imputed, mean)
```
Making our plot
```{r plot3, cache=TRUE, eval=TRUE}
library(lattice)
xyplot(steps ~ interval | weekday, imputedmeans, type = "l", lwd = 1, xlab="Interval", ylab="Steps", layout = c(1,2))
```

